Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-04 paper/paper.tex
undecidability

\begin{abstract}
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs.  To do  this, we analyze programs for
{\em liveness} of heap allocated data, i.e.  data that can possibly be
used in the future.  The accompanying garbage collector is modified to
use the  result of the analysis  to preserve only {\em  live} data---a
subset  of {\em  reachable  data}---during  garbage collection.   This
results  in an  increase  in the  garbage  reclaimed and  consequently
reduces the memory  footprints of programs.  While  this technique has
already been shown to yield  benefits for eager first-order languages,
the presence of  closures at run-time poses  additional challenges for
lazy languages.  Closures require changes both in the earlier liveness
analysis for heap allocated memory  and also in the garbage collection
scheme required to reclaim memory that is not live.

Our  analysis  is a  context-sensitive  analysis  that annotates  each
potential  garbage collection  point  in  the program  with  a set  of
deterministic finite-state  automata (DFA) describing the  liveness at
that  point.   We  also  present a  practical  implementation  of  our
analysis in  a garbage collector.   The collector, when  invoked, uses
the DFA set to preserve live  objects.  As a result, fewer objects are
marked and preserved when compared  to a reachability based marking.
Our experiments  confirm that liveness-based collection  increases the
garbage reclaimed,  consequently decreasing the number  of collections
on many  practical programs.  On  our benchmark programs, we  obtain a
decrease of  up to  y\% in  the number  of collections,  and up  to x\%
decrease in overall garbage collection time.
\end{abstract}

@@

>>>>

\begin{abstract}
We consider  the problem of reducing  the memory required  to run lazy
first-order functional  programs. To do this, we  analyze programs for
liveness of  heap-allocated data.  The  accompanying garbage collector
is modified  to use the result  of the analysis to  preserve only live
data---a subset  of reachable data---during  garbage collection.  This
results  in an  increase  in the  garbage  reclaimed and  consequently
reduces  the memory footprint  of programs.  While this  technique has
already been shown to  yield benefits for eager first-order languages,
the lack of a statically determinable execution order and the presence
of  closures   at  run-time  poses  additional   challenges  for  lazy
languages.   This  requires  changes  both  in  the  earlier  liveness
analysis and also in the garbage collection scheme required to reclaim
memory that is not live.

The implementation  of our  analysis annotates each  potential garbage
collection  point   in  the  program  with  a   set  of  deterministic
finite-state automata (DFA) describing  the liveness at that point. We
also implement a copying collector  that uses the DFA to preserve live
objects.   Our  experiments  confirm  that  liveness-based  collection
always  increases the garbage  reclaimed, consequently  decreasing the
number of collections.  In addition,  it reduces the execution time of
several programs.  Experiments also reveal  that the best  results are
obtained  for   a  mixed  garbage   collection  scheme  that   does  a
liveness-based    collection   for    fully    evaluated   data    and
reachability-based   collection  for   closures.   On   our  benchmark
programs,  we  obtain a  decrease  of  up to  50\%  in  the number  of
collections, and up to 70\% decrease in overall program execution time
when compared to reachability based collectors.
\end{abstract}

@@

\begin{figure*}[t!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e)\!:\!S, H, \kappa$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$ }    &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)$ \\ $\rightsquigarrow  \rho', S, H[z := (H(\rho(x)),H(\rho(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (d_1, d_2)$ & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (\CAR~x)$ \\ $ \rightsquigarrow \rho', S, H[\rho(z) := d_1], e$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S, H, (\CAR~x)$ \\ $\rightsquigarrow \rho', (\rho, x, (\CAR~x))\!:\!S, H, s$}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)$ \\  $\rightsquigarrow \rho', S, H[\rho(z) \mapsto H(\rho(x)) + H(\rho(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S, H, (+~x~y)$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y))\!:\!S, H, s$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho, S, H, (+~x~y)$ \\ $\rightsquigarrow \rho', (\rho, x, (+~x~y))\!:\!S, H, s$}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell{$\rho, S, H, (f~\myvec{x})$ \\ $\rightsquigarrow [\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow s~\IN~e)$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\SIF~x~e_1~e_2), ep, \sigma$ \\ $\rightsquigarrow \rho, S, H,  e_1, ep, \sigma$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\SIF~x~e_1~e_2), ep, \sigma$ \\   $\rightsquigarrow
\rho, S, H,  e_2, ep, \sigma$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,  (\SIF~x~e_1~e_2), ep , \sigma$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2), ep, \sigma)\!:\!S, H,  s, ep_{if}, \sigma$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho, (\rho', z, e, ep, \sigma)\!:\!S, H, (\SRETURN~x), ep_{return}, \sigma$ \\ $\rightsquigarrow \rho', S, H[\rho(z) \mapsto v], e, ep, \sigma$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H, (\SRETURN~x), ep, \sigma$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), ep, \sigma)\!:\!S, H,  s, ep_{return}, \sigma$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}
